Supplementary information - The Functional Therapeutic Chemical Classification System

### 2013-08-22 document.tex
final polishing done
Signed-off-by: loopasam <samuel.croset@gmail.com>

\title[The Functional Therapeutic Chemical Classification System]{The Functional Therapeutic Chemical Classification System}
\author[Croset \textit{et~al}]{Samuel Croset\,$^{1, \footnote{to whom correspondence should be addressed}}$, John Overington\,$^{1}$ and Dietrich Rebholz-Schuhmann\,$^{1}$
}
\address{$^{1}$European Molecular Biology Laboratory
European Bioinformatics Institute (EMBL-EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge CB10 1SD,
United Kingdom\\}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

\maketitle

\begin{abstract}

\section{Motivation:}
Drug repurposing is the discovery of new indications for compounds already approved and used in a clinical setting. 
Recently, some computational approaches have been developed to unveil new opportunities in a systematic fashion, by taking 
into consideration gene expression signatures or chemical features for instance. 
We present here a novel method based on knowledge integration using semantic technologies, 
to capture the functional role of approuved chemical compounds.
\section{Results:}
In order to computationally generate repurposing hypotheses, we used the Web Ontology Language (OWL) to 
mathematically define the meaning of over 20'000 terms formally describing various modes and mechanisms of action. Based on an 
integration of public data, we have automatically assigned over a thousand of approved drugs into 
these MoA categories. The resulting new resource is called the Functional Therapeutic Chemical Classification System (FTC) 
and was further evaluated against the content of the traditional Anatomical Therapeutic Chemical Classification System (ATC). 
We illustrate how the new classification can be used to generate drug repurposing hypotheses, using Alzheimer’s disease as a use-case.

\section{Availability:} A web application built on the top of the resource is freely available at {{https://www.ebi.ac.uk/chembl/ftc}}. The 
source code of the project is available at {{https://github.com/loopasam/ftc}}.

\section{Supplementary Information:}
Supplementary data are available at Bioinformatics online.

\section{Contact:} \href{croset@ebi.ac.uk}{croset@ebi.ac.uk}
\end{abstract}

\section{Motivation}

Drug repurposing is the use of known active compounds for new therapeutic indications \citep{Sanseau01072011}. 
When administered in a living organism, a compound can indeed play various roles and affect different biological processes. 
Accurately identifying these different functions helps to predict the potential side-effects a drug could have and can also lead to 
interesting repurposing opportunities \citep{Medina-Franco2013}. For instance, \emph{sildenafil} was initially developed 
against angina and has been repurposed towards erectile dysfunction during the clinical trials \citep{Ashburn2004}. Approved compounds are 
attractive because they have been extensively studied and have by definition already successfully passed clinical trials, where 
most drugs fail because of safety or efficacy issues.
There is an increasing number of approaches to predict repurposing opportunities using computational methods \citep{Sanseau01072011}. Most 
methods operate on the profiles of physicochemical descriptors derived from molecular structures \citep{Haupt2011}. Other methods characterise 
the drugs on more abstract levels, such as the gene expression signature \citep{Iorio2010} or 
via the reported side-effects \citep{Campillos2008}. 
These approaches 
have in common to look for similarities within existing drugs and forward similar compounds as repurposing hypotheses.
A feature of particular interest to describe drugs is the Mode of Action (MoA). According to Wikipedia, 
the MoA describes \emph{a functional or anatomical change, 
at the cellular level, resulting from the exposure of a living organism to a substance}. For instance 
terms such as \emph{transcriptional regulation agent} or \emph{anticoagulant} define MoAs and characterise the 
roles of a certain type of drugs. The 
MoA abstracts over the relations between molecular functions, protein targets and drug activities; it is the central concept linking a 
chemical structure to a set of biological effects. Intuitively, the indication of a drug logically depends on its MoA.
Despite its widespread use in drug discovery, the MoA has not been used yet as a descriptor for repurposing analyses. One reason for this 
might be the challenge of formally defining MoAs. Indeed, MoAs are terms or categories, it is not possible to represent 
them straightforwardly with values and numbers like one can do for a 3D molecular structure or for a gene expression profile. 
Nonetheless, the meaning of a concept can be formalised with controlled vocabularies and ontologies \citep{Gruber1995}; 
originating from description logics, such frameworks help to define the mathematical meaning of symbols and strings of characters. 
In an ontology or knowledge base, concepts are organised and linked following the logical type of relation they have between them. 
In the Gene Ontology (GO) for example \citep{Ashburner2000}, biological processes and molecular functions terms are manually curated  
and their meaning is specified by the relation types linking two GO terms.
MoA definitions are present in other classifications such as the Medical Subject Headings \citep{Nelson2004} or the Chemical Entities of 
Biological Interest \citep{Hastings2012} for example. The Anatomical Therapeutic 
Chemical Classification System (ATC) \citep{world2000anatomical} also describes to some 
extent the action of drugs at the anatomical level. All these resources are valuable for the community as a source of carefully 
manually curated information; moreover, the categories described in these classification systems are sometimes used to annotate 
drugs: For instance the compound \emph{sildenafil} has been manually annotated as \emph{vasodilator agent} (CHEBI:35620 or MeSH:D27.505.954.411.918).
The classifications mentioned previously are not specially designed for drug repurposing; they purposefully report only the 
well-known and major MoAs of chemical compounds. The pharmacological spectrum of each drug is not necessarily well covered, yet 
it would be the best way to predict new indications. In our context, an ideal knowledge base would feature the known MoAs of a 
drug as well as some predicted ones to be tested in experiments. The MoA categories should derive and scale over primary molecular 
evidences exposed in biomedical databases in an automated way.
To address the lack of systematic MoA annotations, we have implemented the Functional Therapeutic Chemical 
Classification System (FTC), presented here in this manuscript. The FTC is automatically built by leveraging the content 
of various biomedical databases using description logics and automated reasoning. Over 20'000 new MoA categories are defined 
in the resource and further populated with approved drugs using the Web Ontology Language (OWL) in combination with a reasoner. 
Most of the drugs are present in multiple FTC categories, reflecting the various roles a compound can play inside a biological system which 
can serve as starting point for repurposing. 
The resource was evaluated against the ATC, traditional classification scheme introduced before. We present as well some preliminary 
analyses over the data, by looking at the relation between the MoA and the indication of a compound using semantic similarity. 
Finally, we illustrate also how the FTC can be used as a pharmacological toolbox to analyse drug repurposing for Alzheimer’s disease.

\section{Results}

The knowledge base behind the FTC is built by integrating information coming from various sources. 
The GO terms serve as template to create the FTC categories describing the MoAs; DrugBank \citep{Knox2011} provides the 
known links between drugs and their protein targets and Uniprot \citep{TheUniprotConsortium2013} maps targets to their 
respective GO annotations \citep{Dimmer2012}. 
Drugs are further assigned into MoA categories based of the OWL constructs and axioms defined in the 
FTC (see \emph{Availability and Implementation}). A reasoner, a program capable of understanding such axioms, performed this task. 
The process to build the FTC is summarised in Figure~\ref{fig:01}.
It takes around 5 seconds on a modern 
laptop (4 cores processor with 4 Gb of memory) to classify the knowledge base using the ELK reasoner \citep{Kazakov2011}. 
Other OWL reasoners have been tried (e.g, Hermit, Pellet, etc.) but none of them were suitable for this purpose; 
the classification time was always superior to a few minutes (data not shown). The FTC has a taxonomic structure as illustrated 
on Figure~\ref{fig:02}, which arises when the reasoner classifies the knowledge base. Categories can have multiple parents and multiple children. 
The reader can browse and access the content of the FTC online at {{https://www.ebi.ac.uk/chembl/ftc}}.
In total there are 1280 approved DrugBank compounds (chemical and biotherapeutics) perturbing 1264 human protein targets. 
The FTC introduces 23'353 new categories describing the mode and mechanism of action of therapeutic compounds. 4289 of these 
categories are related to biological processes and 19'064 to the molecular functions. A summary of the metrics behind the latest 
build is available online at {{https://www.ebi.ac.uk/chembl/ftc/evaluation/}}.
Out of all FTC classes, 1432 
categories (\textgreater 6\%) directly contain at least one approved drug. This number increases up to 2532 (\textgreater 11\%) 
when direct and 
indirect drugs are considered. FTC categories not containing drugs (e.g, FTC\_A0001771 - \emph{Anti-immunological synapse 
formation agent}) represent modes and mechanisms of action for which no approved compounds exist yet or that have not been identified 
as such in the FTC.

\begin{figure*}[!tpb]%figure1
\centerline{\includegraphics{fig1.png}}
\caption{Summary of the methodology used to construct the resource. (A) The FTC categories are first created from the existing GO terms
describing molecular functions and biological processes. (B) Then, each category has a logical equivalent definition assign to it.
(C) The content of various databases in incorporated and linked using the FTC specific properties. (D) Finally a reasoner runs over the knowledge
base and assigns drugs to MoA categories based on whether or not a definition can be satisfied. The taxonomic structure of the
FTC appears also in the reasoning step, from the entailements of the equivalent defintions.}\label{fig:01}
\end{figure*}

\begin{figure}[!tpb]%figure2
\centerline{\includegraphics{fig2.png}}
\caption{Parent categories to the FTC class \emph{Pro-fibrinolysis agent} (FTC\_P0042730). The classification is a direct 
acyclic graph where categories are describing increasingly specific concepts. Arrows entails subclass relationships 
between the terms (\emph{is a} relation).}\label{fig:02}
\end{figure}

\subsection{Evaluation}
The content of the FTC has been evaluated against the information present in the ATC. The motivation behind 
this step is to have an estimation of how the automatically-built classification (FTC) compares to a manually curated 
one (ATC). In the context of this work, we have considered the ATC as a gold standard against which the FTC 
has been assessed. In practice, these two resources have different goals; therefore the evaluation should be 
interpreted accordingly. The full methodology behind the evaluation is 
described in the section 6 of the supplementary material (\emph{Evaluation methodology}). 
Briefly, 68 categories present in the FTC have an equivalent meaning to a combination of categories present in the ATC. 
This semantic equivalence was manually asserted 
and called \emph{evaluation point} (see supplementary material section 6.1 for details). 
The set of drugs present in both classes for each evaluation points were then compared and some metrics derived. 
For example, the FTC category \emph{Anti-hydrogen:potassium-exchanging ATPase activity agent} (FTC\_A0008900) has 
been manually asserted as equivalent to the ATC category \emph{Proton pump inhibitors} (A02BC). A summary of this 
evaluation point is furthermore available online at {{https://www.ebi.ac.uk/chembl/ftc/evaluation/FTC\_A0008900}}.
Out of the 1280 DrugBank compounds present in the FTC, 1134 are also present in the ATC, therefore only those were 
considered. The \emph{evaluation points} cover a total of 471 DrugBank compounds or around 41\% of common drugs to both classifications.
Out of these, 277 compounds are true positives, 
meaning that given an evaluation point, they are present in both FTC and ATC categories. The \emph{proton pump inhibitor} 
evaluation point is such an example where all the drugs (\emph{omeprazole, esomeprazole, pantoprazole, lansoprazole, rabeprazole}) are 
present both in the FTC category and in the corresponding ATC class(es). The total number of compounds present in an ATC class but 
not in the corresponding FTC category (false negatives) is 36. Finally, 279 compounds are present in a FTC class but not in the 
corresponding ATC class(es); such agents are the false positives. From these values, it is possible to derive a 
recall of 88\%; this percentage indicates that the automatic build of the FTC covers a good portion of the content 
already present in the ATC. The precision of 50\% shows that the FTC contains for a given MoA many more drugs than the 
equivalent ATC categories. This result was expected and comes the original idea behind the FTC: Representing in a systematic fashion
the implicit and explicit MoAs of drugs, in particular the ones not already indexed by current classification scheme.

\subsection{Exploration}
The FTC was designed to assist drug repurposing analyses by explicitly representing the polypharmacology of 
approved drugs. In this section we exemplify how the resource can be used to perform different types of analysis.

\subsubsection{Polypharmacology spectrum}
With more information on a drug’s molecular targets and their physiological roles, the more opportunities 
exist to re-orient a drug into doing something new. The therapeutic agents described in the FTC can have several MoAs, 
representing the intrinsic polypharmacology of the approved compounds. Figure \ref{fig:03} illustrates the polypharmacology 
spectrum by showing the distribution of number of MoAs per compound. When only direct categories are considered, compounds belong on average
to 13.5 MoA categories. This number increases to 61.2 when parent categories are taken into consideration (super classes).
Not all the MoAs are disease relevant, 
some FTC categories are particularly abstract (e.g, \emph{Anti-biological process agent}) yet they represent discrete 
categories to which the drug belongs with an explicit and clear meaning. These discrete MoAs are a good starting 
point to understand what a compound can do when administered in a human system, as it will be shown with Alzheimer's disease.
The human readable definitions 
associated to each FTC category furthermore help the user to choose the adequate MoA for the desired outcome.
According to these results, compound's polypharmacology is 
well represented in the FTC, as shown by the numerous MoAs each approuved drug can exhibit.

\begin{figure*}[!tpb]%figure3
\centerline{\includegraphics{fig3.png}}
\caption{Distribution of the direct (blue) and indirect (red) number of MoAs per drug.
Means are indicated with a solid line. On average each compound has 13.5 MoAs when only direct classes are considered. The number rises
to 61.2 when indirect MoAs are included. Indirect MoAs are the ancestor classes in the taxonomy as shown in Figure \ref{fig:02}.
The distribution range is wider when indirect MoAs are considered (range=299; min=7; max=306) versus direct MoAs only (range=79; min=3; max=82).
These results emphasises the fact that some drugs
are well characterised in databases and could be used for a variety of specific biological tasks. Finally some compounds have been assigned
to a small number of FTC categories; in such cases little is known or reported about their pharmacology and repurposing opportunities
might be limited.}\label{fig:03}
\end{figure*}

\subsubsection{Drugs with similar MoAs have similar indications}
The list of MoAs attributed to a drug can be exploited as a descriptor for the therapeutic agent:
The tree structure of the FTC can be used to derive some similarity metrics over the MoAs. The underlying 
heuristic is to assume that the closer two entities are in the taxonomy, the more similar they are. 
We used a straightforward approach derived from the Jaccard index (see supplementary section 7) in order to compare 
approved drugs based on the similarity of their MoAs. For instance, the similarity between two compounds present in 
the same FTC category is 1 (maximum). The similarity between an \emph{anti-blood coagulant} and \emph{pro-blood coagulant} is 0.29, 
reflecting the fact that such compounds are not similar in regards to the outcome of their biological effect. As the MoA is 
intuitively expected to be the central concept leading to the indication of the drug, we expected that on average, drugs with 
similar MoAs would be indicated towards similar therapeutic areas. The heat map presented in Figure \ref{fig:04} shows a pairwise 
comparison of all the drugs of the FTC based on their relative MoA similarity. The compounds are further grouped by 
therapeutic indications as defined by the ATC.
The heatmap (Figure \ref{fig:04}) reveals some square patches around the central diagonal;
the overall similarity appears higher when compounds from the same ATC group are considered. 
A significance analysis (details covered in supplementary section 8)
revealed that the average MoA similarity of compounds belonging to the same ATC category is 
significantly higher than when compounds belonging to different categories are compared. Indeed, for 
each category, the p-value was inferior to 0.05 based on 20'000 random permutations over the similarity values. 
This result supports the idea that drugs with similar MoAs have similar indications. Note that the mean of the 
similarity values was considered for the statistical analysis; some outliers are also present in the map, which 
can be interpreted as repurposing hypotheses. These outliers have indeed similar MoAs, yet they belong to totally 
different therapeutic areas and are used for different purposes according to the ATC. We are currently further analysing 
such cases in a systematic fashion. Supplementary Figure 3 present similar association behavior when two levels of the ATC are 
considered (no statistical significance performed). Supplementary Figure 4 re-uses the same data as Figure \ref{fig:04} (one ATC level) 
but with a clustering function apply to it (hierarchical clustering - manhattan distance) in order to reveal 
functional clusters of drugs. Finally, Supplementary Figure 5 shows the distribution when compounds are sorted based on 
their identifiers; no patterns are identifiable in this case. Taken together, these results emphasize that the MoAs as 
defined in the FTC are indeed on average associated with the therapeutic indication of a drug. This result supports the validity of the
resource and its potential to computationally address indication discovery.

\begin{figure*}[!tpb]%figure4
\centerline{\includegraphics{fig4.png}}
\caption{Pairwise comparison of MoAs similarities. Only the first level of 
the ATC is considered. The reader can refer to Supplementary Figure 1A for a two ATC level granularity. 
The similarity descriptor ranges from 0 (not similar - white) to 1 (identical - black). Drugs are grouped according 
to their first ATC level (colours on the side). For instance, the compound reteplase (DB00015) has the ATC code B01AD07, 
which appears as ``B'' (dark orange) on the plot.
Some compounds belong to multiple ATC categories (Mulitple - black) and some others do not have
an ATC code (NoCategory - white)
The average similarity of drugs present in the same therapeutic category 
is significantly higher on average when separately compared to all other indications.}\label{fig:04}
\end{figure*}

\subsection{Drug repurposing hypotheses for Alzheimer's disease}
In this section we provide examples how actual drug repurposing hypotheses can be derived from the FTC. 
The approach presented here makes use of the FTC categories as analogs to compartments of a toolbox helping to 
find drugs to treat Alzheimer's disease. Five FTC categories containing drugs are directly related to the biological 
processes of the neurodegenerative condition: \emph{Anti-amyloid precursor protein biosynthetic process agent} (FTC\_A0042983), 
\emph{Anti-tau protein kinase activity agent} (FTC\_A0050321), \emph{Anti-tau protein binding agent} (FTC\_A0048156), 
\emph{Anti-beta amyloid 
binding agent} (FTC\_A0001540) and finally \emph{Pro-beta amyloid binding agent} (FTC\_P0001540). We have then considered the drugs present 
inside each of these classes as potential candidates. Figure \ref{fig:05} shows these drugs, which have been further manually grouped 
based on the overall similarity of their actions (numbers on Figure \ref{fig:05}).

>>>>

\title[Supplementary information - The Functional Therapeutic Chemical Classification System]{Supplementary information - 
The Functional Therapeutic Chemical Classification System}

\section*{Availability and Implementation}

This section describes the building mechanism behind the FTC. 
First, a list of categories describing the mode and mechanism of action of drugs is defined. 
Then in a second step the newly created categories are automatically populated with approved compounds. 
Finally, the FTC is evaluated and repurposing hypotheses can be generated.

\section{Source code}
The code behind the creation of the resource is entirely open and available 
at {{https://github.com/loopasam/ftc}}. The web application built on the top of the 
FTC can be find at {{https://www.ebi.ac.uk/chembl/ftc}} and the documentation can be 
accessed at {{https://github.com/loopasam/ftc/wiki}}. The reader should be familiar with 
description logics and the Web Ontology Language (OWL) to fully understand the construction of the 
knowledge base. An introduction to description logics from the perspective of the biomedical scientist is 
available on the wiki at {{https://github.com/loopasam/ftc/wiki/Description-Logics}}. The FTC implementation 
relies mostly on Brain \citep{Croset2013} and the web application builds on the top of the Play! framework. 
Classification tasks use the ELK reasoner \citep{Kazakov2011}. The computer hosting the web application has 8 Gb of memory 
with 4 processors, this architecture allows fast parallel reasoning, thanks to ELK's design. More functionalities 
will be added to the web application following user requirements (so called \emph{lean implementation}).

\section{FTC categories creation}
The mode of action categories present in the FTC are defined based on the terms coming from the 
Gene Ontology (GO) \citep{Ashburner2000}. Both the molecular function and biological process branches are used for 
this purpose, yet handled slightly differently.

\subsection{Categories related to biological processes}\label{catprocess}
All the biological processes featured by the GO are looked-up one by one. All the 
time a process is linked to another process (\emph{X}) via a \emph{positive} or \emph{negative regulation} link, two FTC 
classes are created: \emph{Anti-X agent} and \emph{Pro-X agent}. For instance the GO term \emph{positive regulation of blood 
coagulation} is linked to the term \emph{blood coagulation} via a \emph{positively regulates} relation, therefore two FTC 
categories \emph{Anti-blood coagulation agent} and \emph{Pro-blood coagulation agent} are created. The identifiers of the 
new FTC classes are also derived from the GO term used to create the class pattern. The GO numeric identifier 
is re-used and the letter \emph{A} or \emph{P} is appended before to emphasize the \emph{anti} or \emph{pro} pattern. 
From the example presented 
previously, the FTC class \emph{Anti-blood coagulation} has FTC\_A0007596 as identifier, because the GO term \emph{blood coagulation} is 
referenced by GO:0007596. Following the same logic, FTC\_P0007596 is the identifier of the class \emph{Pro-blood coagulation agent}. The 
design choice for identifiers and labels allows the FTC to fully rely on the high quality work provided by the GO curation 
team and scale over it.

\subsection{Categories related to molecular functions}\label{catfunc}
The mode and mechanism of actions related to molecular functions are created in the following manner: All the time 
a molecular function (\emph{Y}) is encountered then two FTC categories are created, as for the processes: \emph{Anti-Y agent} and 
\emph{Pro-Y agent}. The identifiers are assigned the same way as described before. For instance, out of the GO term \emph{androgen 
receptor activity} (GO:0004882) two FTC classes are derived: \emph{Pro-androgen receptor activity agent} (FTC\_P0004882) and 
\emph{Anti-androgen receptor activity agent} (FTC\_A0004882).

\section{Equivalent definitions}
FTC classes are generated as presented in the previous section. Up to this point, these categories 
are only tokens with a human readable label as well as an identifier. The next step is going to 
assign equivalent definitions to each FTC class. A reasoner can understand such definitions and will automatically 
classify the knowledge base accordingly; drugs will then be assigned to FTC categories and the taxonomic structure 
arises after this reasoning step. Equivalent definitions are written as OWL class expressions using the entities of 
the knowledge base (summarised at {{https://github.com/loopasam/ftc/wiki/Knowledge-Base}} and in section \ref{specs}).
There are two types of equivalences: The first one captures perturbation of regulatory 
biological processes (so called \emph{regulatory patterns}) and the second one handles the perturbed functions (\emph{functional patterns}).

\subsection{Regulatory pattern}
Some of the FTC categories are created from the biological processes present in the 
GO (cf section \ref{catprocess}); these categories have two arbitrary equivalent definitions, representing the 
two possible ways a compound might impact the biological process. Anti-biological process agent FTC categories 
contain the drugs that negatively perturb a target involved in the positive regulation of the biological process. 
The \emph{anti} categories also feature the compounds that positively perturb a negative regulator of the same process. 
The \emph{pro} categories are equivalent to the opposite pattern. Supplementary Figure \ref{fig:S01} illustrates the equivalent definitions for the 
FTC class \emph{Anti-blood coagulation agent} (FTC\_A0007596).

\begin{figure}[!tpb]%figureS1
\centerline{\includegraphics{figS1.png}}
\caption{Example of equivalent definitions (gray boxes) for the FTC category \emph{Anti-blood coagulation 
agent} (FTC\_A0007596). After the data integration step, a reasoner will look the definitions and identify 
which drugs satisfy the FTC definitions. Definitions are expressed using the Web Ontology Language (OWL), 
serialised here using the Manchester syntax.}\label{fig:S01}
\end{figure}

\subsection{Functional pattern}
The FTC categories generated from the GO molecular functions (cf section \ref{catfunc}) are also equivalent to a logical 
definition. \emph{Anti} FTC categories dealing with molecular activities are asserted as equals to the drugs 
that negatively perturb the function. \emph{Pro} categories are equivalent to the drugs that positively perturb the 
function of interest. A summary of the patterns definitions is available on the 
online wiki at {{https://github.com/loopasam/ftc/wiki/Mode-of-Action}}.

\section{Data integration}
At this stage, the knowledge base contains the created FTC classes associated with their 
logical definitions, as well as the GO and the core FTC entities. The knowledge base is then 
further populated with some information coming from various public databases. Only manually curated 
information is considered.

\subsection{DrugBank}
The DrugBank database is a unique bioinformatics and cheminformatics resource 
that combines detailed drug data with comprehensive drug target information \citep{Knox2011}. The approved 
drugs (small molecules and biotherapeutics) acting on proteins are extracted from the database and 
imported in the FTC knowledge base. In order to be selected, a compound must firstly be approved and 
secondly have a pharmacological action on at least one human protein target present in Uniprot \citep{TheUniprotConsortium2013}. 
The protein targets all have at least one manually asserted GO annotation \citep{Dimmer2012} for a biological process or a 
molecular function. DrugBank links compounds to targets via \emph{actions}. The DrugBank actions are somehow 
structured and consistent: Concepts such as \emph{inhibitor} or \emph{agonist} are reused throughout the database for 
example, yet they are not strictly formalised as a controlled vocabulary. These actions are manually 
standardised to the core properties of the FTC according to their biochemical meaning: For instance the action 
\emph{antagonist} is mapped to the FTC \emph{negatively-perturbs} property.
Compounds coming from DrugBank are represented as OWL classes and asserted 
as subclasses of the class \emph{DrugBank compound} (FTC\_C2). Protein targets are described as OWL classes 
too and subclasses of the core class \emph{Protein}. Each DrugBank compound is then connected to 
its target via the following axiom pattern: \emph{[drug] SubClassOf perturbs some [protein]}. 
E.g. \emph{Ximelagatran SubClassOf negatively-perturbs some Prothrombin}.

\subsection{Gene Ontology annotations}
The GO annotation program aims to provide high-quality GO annotations to 
proteins in UniProt \citep{Dimmer2012}. In the context of the FTC, such annotations are used 
to create axioms linking protein targets to molecular functions and biological processes. 
Each protein annotated with a function creates an axiom such 
as \emph{[protein] SubClassOf has-function some [molecular function]}. 
Each protein annotated to a biological process creates an axiom such 
as \emph{[protein] SubClassOf involved-in some [biological process]}. 
E.g. \emph{Prothrombin SubClassOf involved-in some positive regulation of blood coagulation}. 
Each protein can be involved in multiple processes and capable of realizing multiple functions;
the polypharmacology is captured at this level.

\section{Knowledge base classification}
The knowledge base is fully built at this step and contains core classes, mode 
of actions descriptions alongside the actions of approved DrugBank compounds on 
protein targets in Uniprot. The proteins are linked to their molecular functions and involvement in 
biological processes via the GO annotations. The logical specifications of the FTC are there to glue 
the different data together and to explicitly express the logical links between resources. The FTC 
knowledge base follows an OWL2 EL profile \citep{Motik2009}, which enable the use of fast and parallelised reasoners such 
as ELK. During the classification process, the reasoner checks whether the mode of action equivalent 
definitions are satisfied or not and assigns drugs inside the corresponding FTC categories. The tree structure 
of FTC appears also at this step from the logical definitions.

\section{Evaluation methodology}
As the classification of therapeutic agents is done in an automated way, it 
is important to evaluate the results generated against a known resource which will be 
considered as gold standard. The assessment of the FTC is done against another similar 
classification, the Anatomical Therapeutic Chemical Classification System (ATC) \citep{world2000anatomical}). 
The ATC has been developed to serve as a tool for drug utilization research in order to 
improve quality of drug use. In this resource, the information is manually curated, and drugs 
are assigned to categories based of their legally approved indications. The goal of the ATC differs 
from the one of the FTC, yet the two resources are sharing some very similar concepts, which can be 
used for the evaluation. Categories of both classifications contain approved drugs with a DrugBank 
identifier, meaning that some of the drugs indexed in the FTC are also present in the ATC. From that, 
it is possible to define some evaluation points, which will help to assess the automated classification process.

\subsection{Evaluation Points}
An evaluation point is defined as an equivalence between a class from the FTC with one or 
more classes from the ATC. The idea is to look at the set of drugs contained in both side of the 
equivalence and estimate the overlap, as illustrated in the Supplementary Figure \ref{fig:S02}. Evaluation points are defined 
by hand and not themselves evaluated. The full list of evaluation points as well as a summary of the results 
are available online at https://www.ebi.ac.uk/chembl/ftc/evaluation/. Each evaluation point has a series of 
true/false positive and false negative drugs associated with it.

\begin{figure}[!tpb]%figureS2
\centerline{\includegraphics{figS2.png}}
\caption{Example of evaluation point. Classes from the ATC are manually 
mapped to classes from the FTC. The drugs present in each side of the equivalences 
are compared, and some metrics are derived from it (false positive/negative and true positive). 
A summary of the evaluation is available online at {{https://www.ebi.ac.uk/chembl/ftc/evaluation/}}.}\label{fig:S02}
\end{figure}

\subsection{True Positives}
Drugs that are present in both the FTC and the equivalent ATC class(es) are called true positives. 
These compounds reflect that the automated classification was capable of retrieving correctly the 
information present in the gold standard (ATC).

\subsection{False Negatives}
These drugs are present in the ATC class(es) but not in the corresponding FTC class. 
The automated classification failed to retrieve these compounds. The smaller the number of 
false negatives is, the better the FTC is at recalling drugs. A small number of false negatives 
means that if a drug is present in the ATC (gold standard), then it is likely that the drug will 
also be correctly categorised in the FTC.

\subsection{False Positives}
The false positives are the drugs present in the FTC category of the evaluation point but not in the corresponding ATC classes. 
A high number of false positives means that the FTC is over-assigning compounds to classes.
The false positives relates to the accuracy of the classification. In the context of this work, 
some false positives could also be considered as drug repurposing opportunities.

\subsection{Precision}
Precision is the probability that a randomly selected drug from the FTC is present in the equivalent ATC classes. 
The value is standardised as a percentage and corresponds to the formula: True Positive / (True Positive + False Positive).

\subsection{Recall}
Recall is the probability that a randomly selected drug from the ATC has been assigned to the correct 
corresponding class in the FTC. The value is standardised as a percentage and corresponds 
to the formula: True Positive / (True Positive + False Negative).

\section{Semantic similarity}
The semantic similarity measure performed over the FTC is a derivative of the 
Jaccard index \citep{jaccard1912distribution} \citep{rogers1960computer} (see \cite{pesquita2009semantic} for review). 
It is probably best understood as an example: If we consider two classes A and B, the semantic similarity 
between these classes corresponds to the number of OWL superclasses (direct and indirect, obtained with a reasoner) that 
are shared by A and B (intersection) divided by the number of superclasses of A or B (union). The index ranges 
from 0 (totally different) to 1 (identical).

\section{Mode of action similarity against indication}
A statistical analysis was performed over the data presented on Figure 4. 
When two compounds are randomly taken, they have on average a higher mode of action similarity when 
they are assigned to the same ATC category (one ATC level).
In order to estimate whether this observation was 
due to chance only, we formulated the following null hypothesis (H0): \emph{For a pair of drug A and B, it does not 
matter to which ATC category they belong to, their similarity is always average.}
The alternative hypothesis (H1) was: \emph{For a pair of drug A and B, if A and B have the same ATC code, 
we expect on average to obtain a higher similarity value than if A and B have different ATC codes.}

A permutation test was then performed for each ATC category. For example, we started with the ATC 
category A (first row on Figure 4), looked at the similarity values when pairs of compounds both belong 
to the category A (top right corner square) and compared it to the similarity values when the pair of compounds 
belong to different categories (A and B, A and C and so forth). For each comparison we obtained two distributions of 
values (not shown). On average the similarity values are always higher when the two compounds 
belong to the same category (A/A versus A/B for instance). A permutation test (n = 20'000) was 
performed in order to see whether this observation was due to chance only. We were able to reject the null 
hypothesis for a significance level of 0.05 all the times. 
The choice for a permutation test was driven by the fact that MoA similarity values do not follow any type 
of standard distribution (data not shown). For the significance levels set, the null hypothesis was rejected for 
all first level ATC categories.

\section{Knowledge base specification}\label{specs}
This section presents the scaffold of the knowledge base underlying the FTC. 
The logic structuring the FTC comes essentially from a set of core OWL properties (rich RBox). 
Some of these properties originate from the GO. When necessary some new ones have also been introduced. 
In order to understand how these properties interact, first will be presented the fundamental classes present 
at the top of the FTC classification.

\subsection{Core classes}
The high level concepts covered by the FTC are represented as OWL classes and enumerated below. 
Some of these core classes are coming from external ontologies, in which case the original URI is preserved.

\begin{itemize}
\item \textbf{Molecular Function} \\
Identifier: http://purl.obolibrary.org/obo/GO\_0003674 \\
Label: molecular function \\
Definition: As defined by the Gene Ontology. Elemental activities, such as catalysis or binding, describing 
the actions of a gene product at the molecular level. A given gene product may exhibit one or more molecular functions.
\end{itemize}

\begin{itemize}
\item \textbf{Biological Process} \\
Identifier: http://purl.obolibrary.org/obo/GO\_0008150 \\
Label: biological process \\
Definition: As defined by the Gene Ontology: Any process specifically pertinent to the functioning of 
integrated living units: cells, tissues, organs, and organisms. A process is a collection of molecular 
events with a defined beginning and end.
\end{itemize}

\begin{itemize}
\item \textbf{Protein} \\
Identifier: http://purl.uniprot.org/core/Protein \\
Label: protein \\
Definition: As defined by Uniprot. Description of a protein.
Comment: Gene products present inside the FTC are all human proteins. Uniprot URIs are used.
\end{itemize}

\begin{itemize}
\item \textbf{Drug} \\
Identifier: http://schema.org/Drug \\
Label: drug \\
Definition: As defined by schema.org: A chemical or biologic substance, used as a medical therapy, that has a 
physiological effect on an organism. \\
Comment: In the context of the FTC, DrugBank chemicals are considered for their role as therapeutic agent 
rather than for their chemical structure. Mode of action classes are representing agent types and are subclasses of this concept.
\end{itemize}

\begin{itemize}
\item \textbf{Therapeutic Agent} \\
Identifier: https://www.ebi.ac.uk/chembl/ftc/FTC\_C1 \\
Label: therapeutic agent \\
Definition: Role of a drug capable of producing a therapeutic effect.
\end{itemize}

\begin{itemize}
\item \textbf{DrugBank Compound} \\
Identifier: https://www.ebi.ac.uk/chembl/ftc/FTC\_C2 \\
Label: DrugBank compound \\
Definition: Drug coming from DrugBank.
\end{itemize}

\subsection{Core properties}
The expressivity of the FTC comes mostly for the properties (RBox). Below is a list of the 
OWL object properties present inside the FTC knowledge base.

\begin{itemize}
\item \textbf{Part Of} \\
Identifier: http://purl.obolibrary.org/obo/BFO\_0000050 \\
Characteristic: Transitive \\
Label: part-of \\
Definition: As defined and used in the Gene Ontology.
\end{itemize}

\begin{itemize}
\item \textbf{Has Part} \\
Identifier: http://purl.obolibrary.org/obo/BFO\_0000051 \\
Characteristic: Transitive \\
Label: has-part \\
Definition: As defined and used in the Gene Ontology.
\end{itemize}

\begin{itemize}
\item \textbf{Regulates} \\
Identifier: http://purl.obolibrary.org/obo/RO\_0002211 \\
Chained property: regulates o part-of -\textgreater regulates \\
Label: regulates \\
Definition: As defined and used in the Gene Ontology.
\end{itemize}

\begin{itemize}
\item \textbf{Negatively Regulates} \\
Identifier: http://purl.obolibrary.org/obo/RO\_0002212 \\
SubPropertyOf: regulates \\
Label: negatively-regulates \\
Definition: As defined and used in the Gene Ontology.
\end{itemize}

\begin{itemize}
\item \textbf{Positively Regulates} \\
Identifier: http://purl.obolibrary.org/obo/RO\_0002213 \\
SubPropertyOf: regulates \\
Label: positively-regulates \\
Definition: As defined and used in the Gene Ontology.
\end{itemize}

\begin{itemize}
\item \textbf{Involved In} \\
Identifier: https://www.ebi.ac.uk/chembl/ftc/FTC\_R1 \\
Label: involved-in \\
Domain: protein \\
Range: biological process \\
Definition: Entails the participation of a protein in a biological process. \\
\end{itemize}

\begin{itemize}
\item \textbf{Has Function} \\
Identifier: https://www.ebi.ac.uk/chembl/ftc/FTC\_R2 \\
Label: has-function \\
Domain: protein \\
Range: molecular function \\
Definition: Describes the molecular function a protein can realize. \\
\end{itemize}
